Supplementary Information: Vulnerability of tree growth

### 2025-04-01 doc/manuscript_hkk_drought_sensitivity.Rmd
cleanish intro

<!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
Drought events can threaten this sink at different temporal scales; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
However, the study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
We know that drought characteristics [], species traits, and microenvironment [@bennett_larger_2015;@mcgregor_tree_2021] all affect tree growth responses. 
However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
<!--The above statement is true at least to my knowledge offhand, but we need to check--> 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; for example, a species with low sensitivity to past droughts would be among the least sensitive species in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
Droughts are notoriously hard to define and encompass a variey of types based on the meteorogical/climatic variables affected [@slette_how_2019]. 
Here we define drought as hot/dry conditions anomalously departing from long-term means during any given time of year. 
<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
Tree growth responses to drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020]. 
(Here we distinguish dry season from drought with differences in peridicity; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions.) 
Seasonally dry tropical forests are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. 
Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under periodic hot/dry conditions.  
- evolutionary timescales - e.g. dry season deciduousness   
- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
- Species with drought tolerant traits like deep roots, more negative turgor loss point and ... can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].   
- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Individual growth responses to drought can also depend on the specific microenvironments they experience that can lead to distict atmospheric and hydrological stresses with ecological consequences.  
- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years. 
- ecological timescales - e.g. distribution of trees in relation to wetness  
-  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.  
- Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama]  
- Even when trees are accessing deeper water, this does not mean that they're in better shape during drought. Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].  
- indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought  
- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.  
- Moreover, drought characteristics can positively or negatively interfere with these factors, leading to consistent or heterogenous sensitivities across drought events.

<!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
Given the diversity of factors that can influence drought sensitivity in seasonally dry tropical forests, sampling strategies need to effectively span this variation and have adequate precision.  
- Species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
- Tape measurements can be representative (although often not annual), not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm. (Median annual growth across trees in HKK is ~1.4 mm, deviation from annual growth needs more precision).  
- In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
- Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. -->

 <!-- Within species, size, exposure and location can affect sensitivity to drought.  
- There is theory and evidence that tree size itself matters.  
- Theoretically, we expect that greater height makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium, as described by Darcy's law [@mcdowell_darcy_2015; @fernández-de-uña_role_inpress].
- Height... [see refs in @vinod_thermal_2023; @olson_plant_2018; @couvreurWaterTransportTall2018].
- As tree size increases, leaves exert lower control over hydraulic resistance [@wolfe_leaves_2023].
- @chen_hydraulic_2022  
- There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
- Microclimate buffering leads to cooler, moister understory air [@vinod_thermal_2023].
- Soils under closed canopies would also be cooler during hot times of the year [@lembrechts_global_2022].
- Reduced evaporative demand would also make them moister, and this might be added to by hydraulic redistribution.
- Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->

@@

To characterise these two droughts, we used climate data from a weather station near the plot and publicly available climate data for the region. 
We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
We first calculated daily values of Vapour Pressure Deficit using the equation $$
0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
$$
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
Finally, we calculated the long-term mean for each variable as the mean of the values of the rolling mean across all years of the dendroband measurements, and the standard error around this mean. 
We also used the SPEIbase dataset [REF] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.

@@

To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel]. 
Under this assumption, positive DBH change is scored as 1 and any change <=0 is scored as 0. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak]. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.

@@

All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [REF].
We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
We extracted draws, calculated distributions using functions from the _posterior_ package [REF].

@@

## Drought sensitivity across distinct drought events {-}

@@

## Predictors of species-level variation in drought sensitivity {-}

@@

@@

## Predictors of stem-level variation in drought sensitivity {-}

>>>>

<!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ annually and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
Drought events can threaten this sink; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
However, the study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
We know that drought characteristics [], species traits, and microenvironment [@bennett_larger_2015;@mcgregor_tree_2021] all affect tree growth responses. 
However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
<!--The above statement is true at least to my knowledge offhand, but we need to check--> 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; for example, a species with low sensitivity to past droughts would be among the least sensitive species in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
Droughts are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019]. 
Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. 
<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
Tree growth responses to drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020]. 
(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions beyond the seasonal expectations.) 
Seasonally dry tropical forests are also at thermal and water stress tolerance thresholds for many species, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. 
Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions.

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to maximise growth and survival under periodic hot/dry conditions. 
These strategies can range along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions[@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics. 
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g. some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Individual growth responses to drought can depend on the specific microenvironments that trees experience leading to varied atmospheric and hydrological stresses with distinct ecological consequences. 
Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture, leading to greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments []. 
Consequently, larger trees suffer more growth declines during drought [@bennett_larger_2015]; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years. 
To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water []. 
However, these responses can potentially be different across species or with drought severity.
<!--Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama] -->
Trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought. 
Finally, distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth. 
<!-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.  -->
Drought characteristics may positively or negatively interfere with these evolutionary and physiological strategies, leading to consistent or heterogenous sensitivities across drought events.  
<!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
Given the diversity of factors that can influence drought sensitivity in seasonally dry tropical forests, sampling strategies need to effectively span this variation and have adequate precision.  
- Species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
- Tape measurements can be representative (although often not annual), not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm. (Median annual growth across trees in HKK is ~1.4 mm, deviation from annual growth needs more precision).  
- In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
- Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. --> 
 <!-- Within species, size, exposure and location can affect sensitivity to drought.  
- There is theory and evidence that tree size itself matters.  
- Theoretically, we expect that greater height makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium, as described by Darcy's law [@mcdowell_darcy_2015; @fernández-de-uña_role_inpress].
- Height... [see refs in @vinod_thermal_2023; @olson_plant_2018; @couvreurWaterTransportTall2018].
- As tree size increases, leaves exert lower control over hydraulic resistance [@wolfe_leaves_2023].
- @chen_hydraulic_2022  
- There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
- Microclimate buffering leads to cooler, moister understory air [@vinod_thermal_2023].
- Soils under closed canopies would also be cooler during hot times of the year [@lembrechts_global_2022].
- Reduced evaporative demand would also make them moister, and this might be added to by hydraulic redistribution.
- Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->

@@

To characterise these two droughts, we used climate data from a weather station near the plot and publicly available climate data for the region. 
We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
We first calculated daily values of Vapour Pressure Deficit using the equation $$
0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
$$
We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
Finally, we calculated the long-term mean for each variable as the mean of the values of the rolling mean across all years of the dendroband measurements, and the standard error around this mean. 
We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.

@@

To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
Under this assumption, positive DBH change is scored as 1 and any change <=0 is scored as 0. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015]. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.

@@

All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [@bürkner_brms_2017].
We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
We extracted draws, calculated distributions using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].

@@

## Growth timeseries and Drought sensitivity {-}

@@

## Species effects on drought sensitivity {-}

@@

## Microenvironment effects drought sensitivity {-}

@@
